Syntax Is All You Need: A Universal-Language Approach to Mutant Generation

### 2024-02-16 paper/main.tex
fix typos

Developing a large number of mutation
testing tools, each requiring its own development community, workflow
for mutation, and effort to keep up with language changes, may, however, be
unnecessary.  The core problem of mutation testing,
\emph{mutant generation}, is simply a particular instance of the
problem of \emph{program transformation}.  Program transformation
considers the problem of taking a program, $P$, and producing a
program, $P'$ that is the result of applying some function $f$ to $P$
(i.e., $P' = f(P)$).  In mutant generation, we
consider a family of functions $f$ where each $f_{o,loc}$ is an
instantiation of a \emph{mutation operator} $o$  (as classically, if
often informally, defined in the literature) at a particular program
location $loc$ in $P$.  The inclusion of $loc$ in the specification of
$f$ indicates that the problem is a specialization of general
program transformation in that all classic mutation operators we are
aware of are \emph{local}: they only affect a small portion of a
program, genearlly a contiguous string of characters or tokens.

@@

      \item A comparison of an implementation of mutant generation
       based on this approach with existing tools for four important
       programming languages.  The single-language tools range from
       approximately 3,800 LOC to nearly 60,000 LOC, and, we emphasize, each handle
       one programming language.  Our tool, which provides comparable
       core mutation testing functionality, supports over a dozen
       languages using only about 2,200 LOC of Python and less than 500
       lines for rules defining mutation operators over a dozen
       specific languages (and which can effectively mutate an
       essentially unlimited number of other languages).  Our
       approach enables a smaller tool than any other tool
       we examined (by over 1000 LOC) to support a rich set of
       operators for all languages covered by other tools, plus additional languages.
       \item Far from being limited by its implementation size and
         universality, our tool \emph{generates a larger number of valid,
         non-equivalent mutants} for every language we examined than
         any competing, language-specific tool.  That these additional
         mutants
         are meaningful, rather than obscure, is further supported by
         the fact that our mutation scores are similar to those for
         other tools.

@@

The key insight of this paper is that \emph{most mutation operators
  proposed in the literature do not in fact require parsing of source
  code.}  Consider, for example, one of the most commonly used
mutation operators, replacement of arithmetic operations.  We do not
need to parse a program containing the string {\tt ``x + y''} in order to
replace that string with {\tt ``x - y''}.  Instead, this change is
guaranteed to be effected if we simply search the program, represented
as a string, for all occurrences of  {\tt ``+''}, and produce for each
such occurrence, a mutant where tha character is replaced by {\tt
  ``-''}.  Many languages express arithmetic as infix operators between
expressions; if expressed syntactically, mutations for them can be expressed
once, trivially implementing a large fraction of the classic Mothra~\cite{offutt1996experimental}
mutation operators.  In a \emph{syntactical} (but not semantic) sense, we claim, there is almost a ``universal'' (very
abstract, never implemented) programming language, of which almost all
particular programming languages can be treated as mere dialects.

The obvious objection to this simplistic approach is that unless we
parse the code to identify arithmetic expressions, some of these
replacements will be invalid; the universal language is not actually real!  E.g., in a C++ program we will produce
mutants like {\tt x-+} replacing {\tt x++}.  However, most of these
instances can be avoided by slightly more judicious choice of search
target, e.g., instances of {\tt ``+''} where the preceeding or following
character is not another {\tt ``+''}.  This may still in some cases produce
invalid mutants, but rejecting such mutants is easy, since they will fail
to compile.  Of course, we pay the price of
compiling each mutant at generation
time.

@@

In Python regular expression syntax, this expresses the replacement of
strings with a possibly-empty amount of white space followed by a
character pattern indicating a line of source code, with the same
source code, preserving indentation, turned into a comment.
While much more expressive than simple string replacement, regular
expressions are familliar to most developers.
%In principle,
%development of complex regular expressions for operators can even be
%performed by simply providing examples~\cite{bartoli2014automatic}  or
%describing the operator in natural
%language~\cite{zhong2018generating}.  
%These approaches speak to the potential of string manipulation-based approaches
%for mutation testing, without the need for program parsing or analysis.

@@

Parser parser combinators~\cite{vanTonderPPC} is a recent approach enabling
syntax-transformation for multiple languages, and (we show) an apt choice for
mutation testing operators. This key idea is to replace code constructs
common to many languages (e.g., multi-line code blocks delineated by braces)
with simple declarative patterns. Using parser combinators for coarse syntax
matching avoids the complexity of integrating heterogeneous rewrite tools for
mutation testing (e.g., one for Java, one for C++) and human effort to
implement language-specific mutation operators (e.g., where the user has to
learn to write operators for each language tool). The mechanization and
expressive properties of parser parser combinators is covered in prior
literature~\cite{vanTonderPPC}; in this paper we demonstrate that their application is
uniquely effective for universal source-based mutation testing. In
particular, we
build mutation operators with Comby (\url{https://comby.dev}), the tool that implements parser parser
combinators to declaratively match and rewrite source code syntax.

@@

For the most part we expect users to provide build commands, but do
support automatic compilation and comparison of mutants for Python,
Solidity, Fe, Vyper, Swift, Rust, and Java, including Trivial Compiler
Equivalence~\cite{TCE} checks for redundancy of mutants.  These
default handlers can always be over-ridden if they do not work with a
particular build setup (for Python, the approach almost always works,
but in the case of multi-file projects in other languages, it ususally
requires some manual setup).

@@

One concern with our approach is that every time the
program changes, the cost of invalid mutants must be paid again.
However, the locality of mutants and source changes in fact means this
is seldom required.   Because a mutant is asmall source change, updating a mutant to
reflect even large changes to a source file is nearly cost-free, and if the
original mutant was valid, the new mutant will usually also be valid.
New mutant generation is only required for new lines of code and
modified lines of code.  Our
implementation allows the use of the {\tt git merge-file} algorithm to control the
modification of most mutants, and uses our mechanism for
selective mutant generation to force mutation where
there is a conflict in a ``mutant patch.''

@@

\item{RQ2:}  How does the univeral approach, using regular expressions
  or parser parser combinator rules, compare to existing
  single-language mutation tools, in terms of number of generated
  mutants, mutation scores, equivalent mutants, and other evaluation measures used in the literature?

@@

The mean (and median) percentage of mutants that are valid is higher for
Comby for all languages.  On average, even for a single modestly sized
file, using parser combinator combinators saves 100-600 failed
compilations.

@@

We do not report the time taken for mutation generation and testing.
Dextools uses mutant schema, and PIT performs bytecode mutation, so
both are much faster for compiling mutants, but other tools took similar
amounts of time to compile mutants.   Mutant execution time was also
generally similar, except for tools that produced very few mutants.
We expect that our approach is generally slower for initial mutant
generation, due to a higher number of invalid mutants, but that this
cost is relatively small compared to the time required to compile and
execute valid mutants for almost all projects.  An informal study of
large Java projects by the author of the PIT gradle
plugin~\cite{mutantSlow}
  suggests that even a successful compile of a typical project takes
  somewhere between one-sixth to one-third the time required for
  executing tests, and failed compiles are usually faster than
  successful ones, since invalid mutants almost always fail during
  parsing, and so code generation, optimization, and linking are not performed.

We estimated mutant equivalence when possible by examining a sample of
30 randomly selected unkilled mutants for each file.  For some tools,
human inspection of the individual unkilled mutants was not possible.
However, we were of course able to estimtate the percent of unkilled
mutants that were unkillable for our two modes for all files, and
these numbers can be put in the perspective of rates reported in the
literature.  Reports of equivalent mutant percentages in the
literature sometimes report the percent of \emph{all} mutants
(including killed mutants, which obviously cannot be equivalent) that
are equivalent.  We follow Schuler and Zeller~\cite{EquivMut} in
reporting the percentage of \emph{unkilled} mutants that were found to
be equivalent.  Schuler and Zeller found about 45\% of unkilled
mutants of Java programs to be equivalent.  Other reports of lower
rates, e.g, 9\%~\cite{offutt1997automatically} or 7\%~\cite{TCE}, report the percent of
\emph{all} mutants, and so map to much higher rates of
equivalence over \emph{unkilled} mutants.  We suggest that as a
fraction of unkilled mutants, a rate of 30\% or fewer equivalent
mutants is good, and a rate of 10\% or below is very good.

@@

The ease of defining mutation operators,
plus our interest in generating large numbers of mutants and
prioritizing the most interesting, in order to make mutation testing useful for
projects with extremely high quality test suites, or even using
projects using formal
verification (as in the work of Groce et al.~\cite{groce2018verified}), has encouraged us to take a ``let a thousand flowers bloom''
approach to choosing mutation operators.  For example, our approach
provides many more control-flow mutations than most tools, due to our
observation that test suites sometimes fail to check for loop
execution counts other than one and zero.  Few other tools introduce
{\tt break} and {\tt continue} statements into programs, or transform
one into the other.  These mutants are often equivalent (in fact, they
give rise to a large number of our equivalent mutants reported above), but \emph{also
sometimes expose subtle weaknesses in a high quality test suite}, in our
experience.  In general, we take our philosophy of mutation from the
proposal by Gopinath et al.~\cite{Limits,gopinath2017mutation} that
expanding the set of mutation operations and classes of such
operations, in the long run, is more important than using only a small
set of ``good'' operators, in that the end-goal of mutation testing is
to identify weaknesses in test suites, not simply to estimate a single
score.   The price paid for this
expansiveness in allowing ways to detect weaknessess in testing is a
larger number of mutants to execute, and a slightly larger number of
equivalent mutants.  However, this is not a
fundamental limitation or forced choice.  We propose as future work to
split the operators for each level of the hierarchy into
core operators and a set of ``aggressive'' operators
that includes our full current set of operators.

@@

Using regular expressions, and only the universal rules, our tool
generates 10 valid mutants and 34 invalid mutants of the code (22.73\%
valid mutants).  Comby produces an identical 10 valid mutants, but 65
invalid mutants (13.33\% validity).  Both tools produce a large number
of invalid mutants here due to including rules such as adding {\tt
  break} and {\tt continue} that are sufficiently widespread in
applicability to be included in our ``universal'' set but do not apply
to Haskell; we believe that comby's awareness of Haskell syntax in
this case actually gives it \emph{more} opportunities to apply such
``inappropriate'' mutants.

@@

 Another approach is to have LLMs generate mutants for a given program. We experiment with prompting GPT-4 to generate mutants for a simple Haskell TODO-list management program.
 \cref{lst:haskellmutants} shows the mutants that GPT-4 generateed for this program (formatted as before ==> after). Notice that some of the mutants generated make sense, and 
 resemble bugs a developer might make (replacing 0 with 1 or using the
 wrong data type). However, GPT-4 also generated mutants changing
strings for all outputs to the console. These are not fruitful
mutants; the messages (potentially) could be checked,
 but the arbitrary nature of such strings makes them
 poor mutants (which is why no mutation tools we are aware of
 implement such an operator).

>>>>

Developing a large number of mutation
testing tools, each requiring its own development community, workflow
for mutation, and effort to keep up with language changes, may, however, be
unnecessary.  The core problem of mutation testing,
\emph{mutant generation}, is simply a particular instance of the
problem of \emph{program transformation}.  Program transformation
considers the problem of taking a program, $P$, and producing a
program, $P'$ that is the result of applying some function $f$ to $P$
(i.e., $P' = f(P)$).  In mutant generation, we
consider a family of functions $f$ where each $f_{o,loc}$ is an
instantiation of a \emph{mutation operator} $o$  (as classically, if
often informally, defined in the literature) at a particular program
location $loc$ in $P$.  The inclusion of $loc$ in the specification of
$f$ indicates that the problem is a specialization of general
program transformation in that all classic mutation operators we are
aware of are \emph{local}: they only affect a small portion of a
program, generally a contiguous string of characters or tokens.

@@

      \item A comparison of an implementation of mutant generation
       based on this approach with existing tools for four important
       programming languages.  The single-language tools range from
       approximately 3,800 LOC to nearly 60,000 LOC, and, we emphasize, each handles
       one programming language.  Our tool, which provides comparable
       core mutation testing functionality, supports over a dozen
       languages using only about 2,200 LOC of Python and less than 500
       lines for rules defining mutation operators over a dozen
       specific languages (and which can effectively mutate an
       essentially unlimited number of other languages).  Our
       approach enables a smaller tool than any other tool
       we examined (by over 1000 LOC) to support a rich set of
       operators for all languages covered by other tools, plus additional languages.
       \item Far from being limited by its implementation size and
         universality, our tool \emph{generates a larger number of valid,
         non-equivalent mutants} for every language we examined than
         any competing, language-specific tool.  That these additional
         mutants
         are meaningful, rather than obscure, is further supported by
         the fact that our mutation scores are similar to those for
         other tools.

@@

The key insight of this paper is that \emph{most mutation operators
  proposed in the literature do not in fact require parsing of source
  code.}  Consider, for example, one of the most commonly used
mutation operators, replacement of arithmetic operations.  We do not
need to parse a program containing the string {\tt ``x + y''} in order to
replace that string with {\tt ``x - y''}.  Instead, this change is
guaranteed to be effected if we simply search the program, represented
as a string, for all occurrences of  {\tt ``+''}, and produce for each
such occurrence, a mutant where the character is replaced by {\tt
  ``-''}.  Many languages express arithmetic as infix operators between
expressions; if expressed syntactically, mutations for them can be expressed
once, trivially implementing a large fraction of the classic Mothra~\cite{offutt1996experimental}
mutation operators.  In a \emph{syntactical} (but not semantic) sense, we claim, there is almost a ``universal'' (very
abstract, never implemented) programming language, of which almost all
particular programming languages can be treated as mere dialects.

The obvious objection to this simplistic approach is that unless we
parse the code to identify arithmetic expressions, some of these
replacements will be invalid; the universal language is not actually real!  E.g., in a C++ program we will produce
mutants like {\tt x-+} replacing {\tt x++}.  However, most of these
instances can be avoided by slightly more judicious choice of search
target, e.g., instances of {\tt ``+''} where the preceding or following
character is not another {\tt ``+''}.  This may still in some cases produce
invalid mutants, but rejecting such mutants is easy, since they will fail
to compile.  Of course, we pay the price of
compiling each mutant at generation
time.

@@

In Python regular expression syntax, this expresses the replacement of
strings with a possibly-empty amount of white space followed by a
character pattern indicating a line of source code, with the same
source code, preserving indentation, turned into a comment.
While much more expressive than simple string replacement, regular
expressions are familiar to most developers.
%In principle,
%development of complex regular expressions for operators can even be
%performed by simply providing examples~\cite{bartoli2014automatic}  or
%describing the operator in natural
%language~\cite{zhong2018generating}.  
%These approaches speak to the potential of string manipulation-based approaches
%for mutation testing, without the need for program parsing or analysis.

@@

Parser parser combinators~\cite{vanTonderPPC} is a recent approach enabling
syntax-transformation for multiple languages, and (we show) an apt choice for
mutation testing operators. This key idea is to replace code constructs
common to many languages (e.g., multi-line code blocks delineated by braces)
with simple declarative patterns. Using parser combinators for coarse syntax
matching avoids the complexity of integrating heterogeneous rewrite tools for
mutation testing (e.g., one for Java, one for C++) and human effort to
implement language-specific mutation operators (e.g., where the user has to
learn to write operators for each language tool). The mechanization and
expressive properties of parser parser combinators are covered in prior
literature~\cite{vanTonderPPC}; in this paper we demonstrate that their application is
uniquely effective for universal source-based mutation testing. In
particular, we
build mutation operators with Comby (\url{https://comby.dev}), the tool that implements parser parser
combinators to declaratively match and rewrite source code syntax.

@@

For the most part we expect users to provide build commands, but do
support automatic compilation and comparison of mutants for Python,
Solidity, Fe, Vyper, Swift, Rust, and Java, including Trivial Compiler
Equivalence~\cite{TCE} checks for redundancy of mutants.  These
default handlers can always be over-ridden if they do not work with a
particular build setup (for Python, the approach almost always works,
but in the case of multi-file projects in other languages, it usually
requires some manual setup).

@@

One concern with our approach is that every time the
program changes, the cost of invalid mutants must be paid again.
However, the locality of mutants and source changes in fact means this
is seldom required.   Because a mutant is a small source change, updating a mutant to
reflect even large changes to a source file is nearly cost-free, and if the
original mutant was valid, the new mutant will usually also be valid.
New mutant generation is only required for new lines of code and
modified lines of code.  Our
implementation allows the use of the {\tt git merge-file} algorithm to control the
modification of most mutants, and uses our mechanism for
selective mutant generation to force mutation where
there is a conflict in a ``mutant patch.''

@@

\item{RQ2:}  How does the universal approach, using regular expressions
  or parser parser combinator rules, compare to existing
  single-language mutation tools, in terms of number of generated
  mutants, mutation scores, equivalent mutants, and other evaluation measures used in the literature?

@@

The mean (and median) percentage of mutants that are valid is higher for
Comby for all languages.  On average, even for a single modestly sized
file, using parser parser combinators saves 100-600 failed
compilations.

@@

We do not report the time taken for mutation generation and testing.
Dextool uses mutant schema, and PIT performs bytecode mutation, so
both are much faster for compiling mutants, but other tools took similar
amounts of time to compile mutants.   Mutant execution time was also
generally similar, except for tools that produced very few mutants.
We expect that our approach is generally slower for initial mutant
generation, due to a higher number of invalid mutants, but that this
cost is relatively small compared to the time required to compile and
execute valid mutants for almost all projects.  An informal study of
large Java projects by the author of the PIT gradle
plugin~\cite{mutantSlow}
  suggests that even a successful compile of a typical project takes
  somewhere between one-sixth to one-third the time required for
  executing tests, and failed compiles are usually faster than
  successful ones, since invalid mutants almost always fail during
  parsing, and so code generation, optimization, and linking are not performed.

We estimated mutant equivalence when possible by examining a sample of
30 randomly selected unkilled mutants for each file.  For some tools,
human inspection of the individual unkilled mutants was not possible.
However, we were of course able to estimtate the percent of unkilled
mutants that were equivalent for our two modes for all files, and
these numbers can be put in the perspective of rates reported in the
literature.  Reports of equivalent mutant percentages in the
literature sometimes report the percent of \emph{all} mutants
(including killed mutants, which obviously cannot be equivalent) that
are equivalent.  We follow Schuler and Zeller~\cite{EquivMut} in
reporting the percentage of \emph{unkilled} mutants that were found to
be equivalent.  Schuler and Zeller found about 45\% of unkilled
mutants of Java programs to be equivalent.  Other reports of lower
rates, e.g, 9\%~\cite{offutt1997automatically} or 7\%~\cite{TCE}, report the percent of
\emph{all} mutants, and so map to much higher rates of
equivalence over \emph{unkilled} mutants.  We suggest that as a
fraction of unkilled mutants, a rate of 30\% or fewer equivalent
mutants is good, and a rate of 10\% or below is very good.

@@

The ease of defining mutation operators,
plus our interest in generating large numbers of mutants and
prioritizing the most interesting, in order to make mutation testing useful for
projects with extremely high quality test suites, or even using
projects using formal
verification (as in the work of Groce et al.~\cite{groce2018verified}), has encouraged us to take a ``let a thousand flowers bloom''
approach to choosing mutation operators.  For example, our approach
provides many more control-flow mutations than most tools, due to our
observation that test suites sometimes fail to check for loop
execution counts other than one and zero.  Few other tools introduce
{\tt break} and {\tt continue} statements into programs, or transform
one into the other.  These mutants are often equivalent (in fact, they
give rise to a large number of our equivalent mutants reported above), but \emph{also
sometimes expose subtle weaknesses in a high quality test suite}, in our
experience.  In general, we take our philosophy of mutation from the
proposal by Gopinath et al.~\cite{Limits,gopinath2017mutation} that
expanding the set of mutation operations and classes of such
operations, in the long run, is more important than using only a small
set of ``good'' operators, in that the end-goal of mutation testing is
to identify weaknesses in test suites, not simply to estimate a single
score.   The price paid for this
expansiveness in allowing ways to detect weaknesses in testing is a
larger number of mutants to execute, and a slightly larger number of
equivalent mutants.  However, this is not a
fundamental limitation or forced choice.  We propose as future work to
split the operators for each level of the hierarchy into
core operators and a set of ``aggressive'' operators
that includes our full current set of operators.

@@

Using regular expressions, and only the universal rules, our tool
generates 10 valid mutants and 34 invalid mutants of the code (22.73\%
valid mutants).  Comby produces an identical 10 valid mutants, but 65
invalid mutants (13.33\% validity).  Both tools produce a large number
of invalid mutants here due to including rules such as adding {\tt
  break} and {\tt continue} that are sufficiently widespread in
applicability to be included in our ``universal'' set but do not apply
to Haskell; we believe that Comby's awareness of Haskell syntax in
this case actually gives it \emph{more} opportunities to apply such
``inappropriate'' mutants.

@@

 Another approach is to have LLMs generate mutants for a given program. We experiment with prompting GPT-4 to generate mutants for a simple Haskell TODO-list management program.
 \cref{lst:haskellmutants} shows the mutants that GPT-4 generated for this program (formatted as before ==> after). Notice that some of the mutants generated make sense, and 
 resemble bugs a developer might make (replacing 0 with 1 or using the
 wrong data type). However, GPT-4 also generated mutants changing
strings for all outputs to the console. These are not fruitful
mutants; the messages (potentially) could be checked,
 but the arbitrary nature of such strings makes them
 poor mutants (which is why no mutation tools we are aware of
 implement such an operator).
